A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-12 tex/codatadp.tex
minor

\excludeversion{report}
\includeversion{paper}

@@

@@

\begin{lemma}[Corecursive Singletons]%
\label{lem:corecursive-singletons}%
Let $\delta$ be a corecursive codatatype. The domain interpreting $\delta$ is
either infinite or a singleton. In the latter case, $\delta$ has necessarily a
single constructor, whose arguments have types that are interpreted by
singletons.
\end{lemma}

@@

\begin{lemma}[Termination] \label{lem:t}
All derivation trees are finite.
\end{lemma}
\begin{proof}
Consider a derivation tree with root node $\Ec$.
Let $S$ be the set of terms occurring as the argument of a selector in $\tEc$.
Let $T$ be the set of terms in $\tEc$ of finite datatype type.
For each term $t \in T$,
let $S^0_t$ be the set $\{ t \}$,
for $i \geq 0$ let $S^{i+1}_t =$ $S^i_t \cup \{ s( u ) \mid u : \upsilon \in S^i, \upsilon \in \Data, \upsilon \text{ finite}, s \in \Sel^{\upsilon}  \}$,
and let $S^\ast_t$ be the fixed point of this sequence.
We know this is a finite set for each $t$ since the values of $t$'s type are of finite size.
Let $S^\ast$ be the union of $S^\ast_t$ for all $t \in T$,
and let $\tcEc$ be the set of subterms of $\Ec \cup \{ C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) ) \mid t : \tau \in S \cup S^\ast, C_i \in \Ctr^\tau \}$.
In a derivation tree with root node $\Ec$, 
it can be shown by case analysis on the rules of the calculus that each non-root node $\Fc$ is such that 
$\mathcal{T}(\Fc) \subseteq \tcEc$, and moreover thus contains an equality between two terms from $\tcEc$ not occurring in its parent node.
Thus, the depth of a branch in a derivation tree with root node $\Ec$ is at most $\mid \tcEc \mid^2$,
which is finite since $\tcEc$ is finite.
\qed
\end{proof}

\begin{lemma}[Refutation Soundness] \label{lem:rs}
If there exists a closed derivation tree with root node $\Ec$, then $\Ec$ is unsatisfiable in $\thD$.
\end{lemma}
\begin{proof}
We prove this by structural induction on the (closed) derivation tree with root node $\Ec$.
First, if the derivation tree is an application of \rn{Conflict}, \rn{Unify$_2$}, or \rn{Cyclic},
then $\Ec$ is unsatisfiable in $\thD$.
For \rn{Conflict}, this is a consequence of equality reasoning.
For \rn{Unify$_2$}, this is a consequence of distinctness.
For \rn{Cyclic}, our construction of $\Val$ indicates that the class of values that $t_1$ can take in models of $\Ec$ is infeasible,
and thus $\Ec$ is unsatisfiable.
If the children of $\Ec$ are closed derivation trees 
whose root node is the result of applying the rule \rn{Split} on term $t$ of type $\tau$,
then by the induction hypothesis $\Ec \cup t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$ is unsatisfiable
for each $C_i \in \Ctr^\tau$.
Since by exhaustiveness, all models of $\thD$ entail (exactly) one equality $t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$,
we know $\Ec$ is unsatisfiable in $\thD$.
Otherwise, the child of $\Ec$ is a closed derivation tree 
whose root node is $\Ec \cup t_1 \teq t_2$ obtained by applying one of the rules \rn{Refl}, \rn{Symm}, \rn{Trans}, \rn{Cong}, \rn{Unify$_1$}, \rn{Bisimilar}, or \rn{Singleton}.
In all cases, we have that $\Ec \models_{\thD} t_1 \teq t_2$.
For \rn{Refl}, \rn{Symm}, \rn{Trans}, \rn{Cong}, this is a consequence of equality reasoning.
For \rn{Unify$_1$}, this is a consequence of injectivity.
For \rn{Bisimilar}, our construction of $\Val$ indicates that the value of $t_1$ and $t_2$ are observationally equivalent in all models of $\Ec$.
For \rn{Singleton}, clearly $t_1$ and $t_2$ must have the same value since the cardinality of their type is one.
By the induction hypothesis, we have $\Ec \cup t_1 \teq t_2$ is unsatisfiable in $\thD$, 
and thus $\Ec$ is unsatisfiable in $\thD$.
\qed
\end{proof}

@@

\begin{lemma}[Solution Soundness] \label{lem:ss}
If there exists a derivation tree with root node $\Ec$ containing a saturated node, then $\Ec$ is satisfiable in $\thD$.
\end{lemma}
\begin{proof}
Let $\Fc_0$ be a saturated node in a derivation tree with root node $\Ec$.
We will construct a model $\M$ for a set of equalities $\Fc$ that is equivalent to $\Fc_0$, and where 
all equivalence classes of $\Fc$ contain at least one variable, and all selectors in $\Fc$ are applied to variables only.
The former comes with no loss of generality since new equalities of the form $y \teq t$ for fresh variable $y$ can be added to $\Fc$ without affecting its satisfiability,
The latter also comes with no loss of generality since nested applications of selectors in $\Fc_0$ can be replaced by fresh variables while adding additional equalities to $\Fc$.

>>>>

@@

\newcommand\afterDot{\;} %%% Too little space by default after "Lemma Foo."

@@

\begin{lemma}[Corecursive Singletons]%
\label{lem:corecursive-singletons}%
\afterDot
Let $\delta$ be a corecursive codatatype. The domain interpreting $\delta$ is
either infinite or a singleton. In the latter case, $\delta$ necessarily has a
single constructor, whose arguments have types that are interpreted by
singletons.
\end{lemma}

@@

\begin{lemma}[Termination]%
\label{lem:t}%
\afterDot
All derivation trees are finite.
\end{lemma}
\begin{proof}
Consider a derivation tree with root node $\Ec$.
Let $S$ be the set of terms occurring as the argument of a selector in $\tEc$.
Let $T$ be the set of terms in $\tEc$ of finite datatype type.
For each term $t \in T$,
let $S^0_t$ be the set $\{ t \}$,
for $i \geq 0$ let $S^{i+1}_t =$ $S^i_t \cup \{ s( u ) \mid u : \upsilon \in S^i, \upsilon \in \Data, \upsilon \text{ finite}, s \in \Sel^{\upsilon}  \}$,
and let $S^\ast_t$ be the fixed point of this sequence.
We know this is a finite set for each $t$ since the values of $t$'s type are of finite size.
Let $S^\ast$ be the union of $S^\ast_t$ for all $t \in T$,
and let $\tcEc$ be the set of subterms of $\Ec \cup \{ C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) ) \mid t : \tau \in S \cup S^\ast, C_i \in \Ctr^\tau \}$.
In a derivation tree with root node $\Ec$, 
it can be shown by case analysis on the rules of the calculus that each non-root node $\Fc$ is such that 
$\mathcal{T}(\Fc) \subseteq \tcEc$, and moreover thus contains an equality between two terms from $\tcEc$ not occurring in its parent node.
Thus, the depth of a branch in a derivation tree with root node $\Ec$ is at most $\mid \tcEc \mid^2$,
which is finite since $\tcEc$ is finite.
\qed
\end{proof}

\begin{lemma}[Refutation Soundness]%
\label{lem:rs}%
\afterDot
If there exists a closed derivation tree with root node $\Ec$, then $\Ec$ is unsatisfiable in $\thD$.
\end{lemma}
\begin{proof}
We prove this by structural induction on the (closed) derivation tree with root node $\Ec$.
First, if the derivation tree is an application of \rn{Conflict}, \rn{Unify$_2$}, or \rn{Cyclic},
then $\Ec$ is unsatisfiable in $\thD$.
For \rn{Conflict}, this is a consequence of equality reasoning.
For \rn{Unify$_2$}, this is a consequence of distinctness.
For \rn{Cyclic}, our construction of $\Val$ indicates that the class of values that $t_1$ can take in models of $\Ec$ is infeasible,
and thus $\Ec$ is unsatisfiable.
If the children of $\Ec$ are closed derivation trees 
whose root node is the result of applying the rule \rn{Split} on term $t$ of type $\tau$,
then by the induction hypothesis $\Ec \cup t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$ is unsatisfiable
for each $C_i \in \Ctr^\tau$.
Since by exhaustiveness, all models of $\thD$ entail (exactly) one equality $t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$,
we know $\Ec$ is unsatisfiable in $\thD$.
Otherwise, the child of $\Ec$ is a closed derivation tree 
whose root node is $\Ec \cup t_1 \teq t_2$ obtained by applying one of the rules \rn{Refl}, \rn{Symm}, \rn{Trans}, \rn{Cong}, \rn{Unify$_1$}, \rn{Bisimilar}, or \rn{Singleton}.
In all cases, we have that $\Ec \models_{\thD} t_1 \teq t_2$.
For \rn{Refl}, \rn{Symm}, \rn{Trans}, \rn{Cong}, this is a consequence of equality reasoning.
For \rn{Unify$_1$}, this is a consequence of injectivity.
For \rn{Bisimilar}, our construction of $\Val$ indicates that the value of $t_1$ and $t_2$ are observationally equivalent in all models of $\Ec$.
For \rn{Singleton}, clearly $t_1$ and $t_2$ must have the same value since the cardinality of their type is one.
By the induction hypothesis, we have $\Ec \cup t_1 \teq t_2$ is unsatisfiable in $\thD$, 
and thus $\Ec$ is unsatisfiable in $\thD$.
\qed
\end{proof}

@@

\begin{lemma}[Solution Soundness]%
\label{lem:ss}%
\afterDot
If there exists a derivation tree with root node $\Ec$ containing a saturated node, then $\Ec$ is satisfiable in $\thD$.
\end{lemma}
\begin{proof}
Let $\Fc_0$ be a saturated node in a derivation tree with root node $\Ec$.
We will construct a model $\M$ for a set of equalities $\Fc$ that is equivalent to $\Fc_0$, and where 
all equivalence classes of $\Fc$ contain at least one variable, and all selectors in $\Fc$ are applied to variables only.
The former comes with no loss of generality since new equalities of the form $y \teq t$ for fresh variable $y$ can be added to $\Fc$ without affecting its satisfiability,
The latter also comes with no loss of generality since nested applications of selectors in $\Fc_0$ can be replaced by fresh variables while adding additional equalities to $\Fc$.
